Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-05 ind_files/99_Appendix_D_examples.tex
Minor rephrasing

Going from population numbers $\mathbf{n}$ to densities $\mathbf{x} = \mathbf{n}/K$, we obtain the birth and death rates:
\begin{equation}
\label{nD_example_density_b_d_rates}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= \left(\lambda + \sum\limits_{j=1}^{m}\beta_{ij}x_j\right)x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}\delta_{ij}x_j\right)x_i
\end{aligned}
\end{equation}
Thus, we have
\begin{equation*}
A^{\pm}_{i} = x_i\left(\left(\lambda \pm \mu\right) + \sum\limits_{j=1}^{m}\left(\beta_{ij} \pm \delta_{ij}\right)x_j\right)
\end{equation*}
Defining $r \coloneqq \lambda - \mu$, $\nu \coloneqq \lambda + \mu$, and $T_{ij} \coloneqq \beta_{ij} + \delta_{ij}$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by
\begin{equation}
\label{nD_example_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{W}_t
\end{equation}
where 
\begin{equation*}
\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
and
\begin{equation*}
(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(\nu + \sum\limits_{j=1}^{m}T_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
From \eqref{nD_det_limit}, we see that the deterministic limit is a set of $m$ coupled ODEs given by
\begin{equation}
\label{nD_example_det_limit}
\frac{d x_i}{dt} = x_i\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{equation}
These are precisely the Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and
\begin{equation}
\label{nD_example_fitness}
w_i(\mathbf{x}) = r + \sum\limits_{j=1}^{m}M_{ij}x_j
\end{equation}
If $\mathbf{p}(t) = [p_1(t),\ldots,p_m(t)]^{\mathrm{T}}$ is the frequency vector at time $t$ and $N_K(t) = \sum_i x_i(t)$, then the mean fitness is given by
\begin{align}
\overline{w}(t) &= \sum\limits_{i=1}^{m}w_ip_i\\
&= \sum\limits_{i=1}^{m}\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r + \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{align}
where we have used the fact that $\sum_i p_i = 1$ in the last line. Using \eqref{nD_replicator_mutator} to write down the equations for the frequencies $p_i$, we obtain
\begin{equation}
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i   
\end{equation}
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $M$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. Both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation that projects the dynamics onto an appropriate curve: If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics~\citep{constable_mapping_2017,parsons_dimension_2017}. However, I will not explore such dimensional reduction techniques further in this manuscript, and refer the reader to~\cite{constable_stochastic_2013} and~\cite{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.

>>>>

Going from population numbers $\mathbf{n}$ to densities $\mathbf{x} = \mathbf{n}/K$, we obtain the birth and death rates:
\begin{equation}
\label{nD_example_density_b_d_rates}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= \left(\lambda + \sum\limits_{j=1}^{m}\beta_{ij}x_j\right)x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}\delta_{ij}x_j\right)x_i
\end{aligned}
\end{equation}
Thus, we have
\begin{equation*}
A^{\pm}_{i} = x_i\left(\left(\lambda \pm \mu\right) + \sum\limits_{j=1}^{m}\left(\beta_{ij} \pm \delta_{ij}\right)x_j\right)
\end{equation*}
Defining $r \coloneqq \lambda - \mu$, $\nu \coloneqq \lambda + \mu$, and $T_{ij} \coloneqq \beta_{ij} + \delta_{ij}$, and matching terms with equation \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and
\begin{align}
	w_i(\mathbf{x}) &= r + \sum\limits_{j=1}^{m}M_{ij}x_j\label{nD_example_fitness}\\
	\tau_i(\mathbf{x}) &= \nu + \sum\limits_{j=1}^{m}T_{ij}x_j\label{nD_example_turnover}
\end{align}
From equation \eqref{nD_Ito_SDE}, we see that the `mesoscopic' description of the system is the $m$ dimensional SDE given by
\begin{equation}
\label{nD_example_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{W}_t
\end{equation}
where 
\begin{equation*}
\mathbf{A^-}_i = {(\mathbf{X}_{t})}_iw_i(\mathbf{X}_{t})
\end{equation*}
and
\begin{equation*}
(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i\tau_i(\mathbf{X}_{t})
\end{equation*}
From \eqref{nD_det_limit}, we see that the infinite population limit of our model is a set of $m$ coupled ODEs given by
\begin{equation}
\label{nD_example_det_limit}
\frac{d x_i}{dt} = x_iw_i(\mathbf{x}) = x_i\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{equation}
These are precisely the Lotka-Volterra equations for a system of $m$ species.
If $\mathbf{p}(t) = [p_1(t),\ldots,p_m(t)]^{\mathrm{T}}$ is the frequency vector at time $t$ and $N_K(t) = \sum_i x_i(t)$, then the mean fitness is given by
\begin{align}
\overline{w}(t) &= \sum\limits_{i=1}^{m}w_ip_i\\
&= \sum\limits_{i=1}^{m}\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r + \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{align}
where we have used the fact that $\sum_i p_i = 1$ in the last line. Using \eqref{nD_replicator_mutator} to write down the equations for the frequencies $p_i$, we obtain
\begin{equation}
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i   
\end{equation}
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $M$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. Both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation that projects the dynamics onto an appropriate curve: If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics~\citep{constable_mapping_2017,parsons_dimension_2017}. However, I will not explore such dimensional reduction techniques further in this manuscript, and refer the reader to~\cite{constable_stochastic_2013} and~\cite{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.
